Road traffic sound level estimation from realistic urban sound mixtures by Non-negative Matrix Factorization

### 2018-03-06 paper/gloaguenAppliedNMF.tex
proofread intro

\modulolinenumbers[5]

@@

Acoustic sensor networks are increasingly deployed in cities, and appear more and more as a possible tool to enrich modeled road traffic noise maps through data assimilation techniques. This, or more simply the validation of the modeled maps through measures, requires first being able to isolate from the measured sound mixtures the road traffic sound level. This task is anything but trivial because of the multiple sound sources that overlap within urban sound mixtures. In this paper, a Non-negative Matrix Factorization (NMF) framework is developed to estimate road traffic noise levels within urban sound scenes. A corpus of sound scenes is artificially built imitating real annotated recordings. The realism of the scenes is validated through a perceptual test, forming a protocol that both reproduces the sensor network outputs, and in which the actual occurrence and sound level of each source is known. Three variants of NMF are tested, namely supervised, semi-supervised, and threshold initialized NMF. While the semi-supervised approach is the most appropriated for park environments, threshold initialized NMF, which is developed especially for this research purpose, appears to be the best generic approach, allowing road traffic noise level estimation with average errors of 1.3 dB over the tested corpus of sound scenes.

@@

\begin{keyword}
non-negative matrix factorization \sep urban sound environment \sep road traffic sound level
\end{keyword}

@@

Noise mapping has been recommended as a tool to tackle noise pollution, in response to the growing demand from urban dwellers for a better environment. The enactment of the European Directive 2002/EC/49 makes such maps mandatory to cities over 100 000 inhabitants. Those maps play an important informative role, establishing the distribution of the sound levels all over the cities as well as the estimation of the number of city dwellers exposed to high sound level ($>$ 55 dB(A)) \cite{nugent2014noise}. Road traffic concentrates particular attention as it is the main urban source of noise annoyance. Road traffic noise maps are built from data collection that consist of traffic data collected on the main roads (flow rates, mean speeds and heavy vehicle ratio) and urban geographic data (building heights and location, topology, ground surfaces, etc.). Follows sound emission and sound propagation computations, resulting in the production of the two indicators equivalent A-weighted sound levels, $L_{DEN}$ (\textit{Day-Evening-Night}) and $L_N$ (\textit{Night}) \cite{kephalopoulos2012common}. This procedure also enables drawing up action plans to reduce the noise exposure. Despite their unanimously recognized interest, noise maps suffer some limitations. The computer efficiency required to produce noise maps at the city scale calls simplifications of the numerical tools and the simulation models that both generate uncertainties \cite{van_leeuwen_noise_2015}. Data collection is itself a vector of uncertainty. Moreover, the produced aggregated indicators mask the sound levels evolution due to the traffic variations throughout the day.

Therefore, noise measurements are increasingly used in addition to simulation to describe urban noise environments. Several measurement set-ups have been proposed in the last years, including  mobile measurements with high quality microphones \cite{manvell2004sadmam, can2014measurement}, participative sensing through dedicated smartphone applications \cite{picaut:halshs-01565214, ventura2017evaluation}, or the development of fixed-sensor networks. In this latter case, the sensor networks can be based either on high-quality sensors as in \cite{mietlicki2012innovative, maijala2018environmental}, or low-cost sensors as in the DYNAMAP project \cite{dynamap_2016} or the CENSE project \cite{picaut2017characterization}. The costs and benefits of each protocol are discussed. Mobile and participatory measures increase spatial coverage at low cost, but lack temporal representativeness. Fixed networks are very reliable for measuring sound levels temporal variations, but allow only a small spatial coverage of the network. In addition, the low-cost sensors enable a wider deployment, but at the cost of increased uncertainties, the most extreme example being smartphone applications.

@@

All these measurement protocol enables a priori combining measures and modeling to improve the accuracy of the produced noise maps. Traffic noise maps and measurements were compared on restrictive areas in \cite{lefebvre2017traffic} and \cite{mioduszewski2011noise}. Wei et al. \cite{wei_dynamic_2016} modify the acoustical parameters of the simulation thanks to noise measurements, while  Mallet et al. \cite{ventura2017estimation} call for data assimilation techniques between models and measurements to reduce the uncertainty of the produced noise maps. However, these works make the implicit assumption that the noise measurements consist mainly of road traffic. In the aim to improve road traffic noise maps, the use of measurements has first to deal with the challenge to estimate correctly the road traffic sound level.
Even if road traffic is predominant on many urban areas, urban sound environments are composed of many different overlapping sound sources (passing cars, voices, footsteps, car horn, whistling birds \dots), what makes the task of estimating correctly the traffic sound level within an urban sound mixture not trivial.

Many works have dealt with the detection \cite{luitel2016sound} or the recognition \cite{defreville_automatic_2006} of sound events in environmental sound scenes. In these cases, a usually two-step schemes is followed were audio samples are described with a set of features (Mel Frequency Cepstral Coefficient, MPEG-7 descriptors \dots) and classified them with the help of a classifiers (Gaussian Mixtures Models, Artificial Neural Network \dots) \cite{chu2008environmental, cowling_comparison_2003}. The classifiers is learnt from a learning database and are next applied on a test database to validate the algorithms. 
Dedicated to the traffic, in \cite{socoro_anomalous_2017}, an Anomalous Event Detection, based on MFCC features, is proposed with the specific aim to improve the traffic sound estimation. It is based on the detection of sound events in order to discard them.

An other approach, followed in this paper, is to consider the Blind Source Separation paradigm, see Figure \ref{fig:source_separation}, which consists in the extraction of a specific signal inside a set of mixed signals. From the different existing methods (CASA, ICA), Non-negative Matrix Factorization (NMF) \cite{lee_learning_1999}, seems the most relevant method for monophonic sensor networks. Dealing easily with the overlapping between the sound sources, this method approximes the magnitude spectrogram of a single signal by the product of two non negative matrices. A lot of applications can be found for musical \cite{smaragdis_non-negative_2003, benetos2006musical} and speech \cite{wilson_speech_2008, mysore2011non} contents. Dedicated to sound separation, Immani and Kasa\"i \cite{satoshi_innami_nmf-based_2012} used NMF in a two steps sound separation with the help of time variant gain features. A first study \cite{gloaguen2018Estimation} has been conducted, in which diverse NMF formula, namely the supervised, the semi-supervised, and the threshold initialized NMF, have been applied on a large set of simulated sound scenes that which mixed traffic component with specific urban sounds at calibrated sound levels. The study proved the interest of NMF for urban sound environments and compared the benefits of each approach. However, it has now to face to real urban sound scenes.

In this paper, the NMF framework is applied on a corpus of simulated sound scenes, generated based on annotated urban recordings, and whose realism is validated by a perceptual test. The NMF framework and its different implemented versions are described in section \ref{part:nmf}. Next, the corpus of urban sound scenes is presented in section \ref{part:urban_scene}, from the sound database built-up to its validation through a perceptual test. Finally, in section \ref{part:expProtocol} and \ref{part:results}, the experimental protocol and the results are exposed.

>>>>

\newcommand{\ml}[1]{\textcolor{red}{ML : #1}}
\modulolinenumbers[5]

@@

Experimental acoustic sensor networks are currently tested in large cities, and appear more and more as a useful tool to enrich modeled road traffic noise maps through data assimilation techniques. One challenge is to be able to isolate from the measured sound mixtures acoustic quantities of interest such as the  sound level of road traffic. This task is anything but trivial because of the multiple sound sources that overlap within urban sound mixtures.

In this paper, a Non-negative Matrix Factorization (NMF) framework is developed to estimate road traffic noise levels within urban sound scenes. To evaluate the performance of the proposed approach, a synthetic corpus of sound scenes is designed to cover most common soundscape settings. The realism of the produced scenes is validated through a perceptual test, forming a protocol that both reproduces the sensor network outputs, and in which the actual occurrence and sound level of each source is known.

Several variants of NMF are tested. The threshold initialized NMF appears to be the most relibale approach, allowing road traffic noise level estimation with average errors of 1.3 dB over the tested corpus of sound scenes.

@@

\begin{keyword}
non-negative matrix factorization \sep urban sound environment \sep road traffic sound level estimation
\end{keyword}

@@

In response to the growing demand from urban dwellers for a better environment, noise mapping has been recommended as a tool to tackle noise pollution. The enactment of the European Directive 2002/EC/49 makes such maps mandatory to cities over 100 000 inhabitants. Those maps play an important informative role, establishing the distribution of the sound levels all over the cities as well as the estimation of the number of city dwellers exposed to high sound level ($>$ 55 dB(A)) \cite{nugent2014noise}. Road traffic concentrates particular attention as it is the main urban source of noise annoyance. Road traffic noise maps are typically built from data collection that consist of traffic data collected on the main roads (flow rates, mean speeds and heavy vehicle ratio) and urban geographic data (building heights and location, topology, ground surfaces, etc.). Follows sound emission and sound propagation computational techniques, resulting in the production of the two indicators equivalent A-weighted sound levels, $L_{DEN}$ (\textit{Day-Evening-Night}) and $L_N$ (\textit{Night}) \cite{kephalopoulos2012common}. This procedure also enables drawing up action plans to reduce the noise exposure. Despite their unanimously recognized interest, noise maps suffer some limitations. The computing cost required to produce noise maps at the city scale calls simplifications of the numerical tools and the simulation models that both generate uncertainties \cite{van_leeuwen_noise_2015}. Data collection is itself also a vector of uncertainty. Moreover, the produced aggregated indicators do not model the sound levels evolution due to the traffic variations throughout the day.

Noise measurements are thus increasingly used in addition to simulation to describe urban noise environments. Several measurement set-ups have been proposed in the last years, including  mobile measurements with high quality microphones \cite{manvell2004sadmam, can2014measurement}, participative sensing through dedicated smartphone applications \cite{picaut:halshs-01565214, ventura2017evaluation}, or the development of fixed-sensor networks. In this latter case, the sensor networks can be based either on high-quality sensors as in \cite{mietlicki2012innovative, maijala2018environmental}, or low-cost sensors as in the DYNAMAP project \cite{dynamap_2016} or the CENSE project \cite{picaut2017characterization}. The costs and benefits of each protocol are discussed. Mobile and participatory measures increase spatial coverage at low cost, but lack temporal representativeness. Fixed networks are very reliable for measuring sound levels temporal variations, but allow only a small spatial coverage of the network. In addition, the low-cost sensors enable a wider deployment, but at the cost of increased uncertainties, the most extreme example being smartphone applications.

@@

All these measurement protocols allows the combination of measures and predictions to improve the accuracy of the produced noise maps. Traffic noise maps and measurements were compared on restrictive areas in \cite{lefebvre2017traffic} and \cite{mioduszewski2011noise}. Wei et al. \cite{wei_dynamic_2016} modify the acoustical parameters of the simulation thanks to noise measurements, while  Mallet et al. \cite{ventura2017estimation} call for data assimilation techniques between models and measurements to reduce the uncertainty of the produced noise maps. However, these works make the implicit assumption that the noise measurements consist mainly of road traffic. In the aim to improve road traffic noise maps, the use of measurements has first to deal with the challenge to estimate correctly the road traffic sound level.
Even if road traffic is predominant on many urban areas, urban sound environments are composed of many different overlapping sound sources (passing cars, voices, footsteps, car horn, whistling birds \dots), what makes the task of estimating correctly the traffic sound level within an urban sound mixture not trivial.

Many works have dealt with the detection \cite{luitel2016sound} or the recognition \cite{defreville_automatic_2006} of sound events in environmental sound scenes. In these cases, a two-step scheme is followed where audio samples are described with a set of features (Mel Frequency Cepstral Coefficient, MPEG-7 descriptors \dots) and classified them with the help of a classifiers (Gaussian Mixtures Models, Artificial Neural Network \dots) \cite{chu2008environmental, cowling_comparison_2003}. The classifiers is learnt from a learning database and are next applied on a test database to validate the algorithms.
Dedicated to the traffic, in \cite{socoro_anomalous_2017}, an Anomalous Event Detection, based on MFCC features, is proposed with the specific aim to improve the traffic sound estimation. It is based on the detection of unwanted sound events in order to discard them.

An other approach, followed in this paper, is to consider the blind source separation paradigm which consists in the extraction of a specific signal inside a set of mixed signals, see Figure \ref{fig:source_separation}. From the different existing methods, Non-negative Matrix Factorization (NMF) \cite{lee_learning_1999}, appers to be a relevant method for monophonic sensor networks. Many applications can be found for musical \cite{smaragdis_non-negative_2003, benetos2006musical} and speech \cite{wilson_speech_2008, mysore2011non} contents. Dedicated to sound separation, Immani and Kasa\"i \cite{satoshi_innami_nmf-based_2012} used NMF in a two steps sound separation with the help of time variant gain features. A first study \cite{gloaguen2018Estimation} has been conducted, in which diverse NMF estimation rules are compared, namely the supervised, the semi-supervised, and the threshold initialized NMF, have been applied on a large set of simulated sound scenes that which mixed traffic component with specific urban sounds at calibrated sound levels. The study demonstrated the interest of NMF for urban sound environments and compared the benefits of each approach on a validation corpus, designed to . However, it has now to face to real urban sound scenes.

\ml{la fin de ce par est a revoir car c'est le point crucial du papier. Bien expliquer le corpus du papier d'avant son usage et en quoi on passe ici a l'étape superieure.}

In this paper, the NMF framework is applied on a corpus of simulated sound scenes, generated based on annotated urban recordings, and whose realism is validated by a perceptual test. The NMF framework and its different optimizatio strategies are described in section \ref{part:nmf}. Next, the corpus of urban sound scenes is presented in section \ref{part:urban_scene}, from the sound database built-up to its validation through a perceptual test. The experimental protocol and the results are then presented and discussed in section \ref{part:expProtocol} and \ref{part:results}.
